---
title: "Hands on session 4: MLE and Regression"
author: "Ferran Muiños"
date: "11/28/2020"
output:
  BiocStyle::html_document:
      toc: true
      toc_float: true
      number_sections: true
      code_folding: hide
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1400)
```


# Requirements

## Packages

- [dslabs](https://cran.r-project.org/web/packages/dslabs/dslabs.pdf)
- [rootSolve](https://cran.r-project.org/web/packages/rootSolve/vignettes/rootSolve.pdf)


# Introduction and Objectives

The learning objectives of this document are:

1. Understand the statistical assumptions behind regression models
2. Fitting generalized linear models (GLM) in R
3. Hands-on optimization of multivariable functions


# Key concepts


## Random processes and likelihood

- Random processes
- Likelihood formula for independent-identically distributed (i.i.d.) data 
generated by a random process
- Connection between optimization and GLM

## Multivariate optimization

- Looking for critical points
- Classification of critical points
- Quadratic approximation

## R skills

- Review linear algebra tools
- More examples of plots and graphical tools
- Defining functions in R
- Computing the gradient numerically
- Solving numerically arbitrary systems of equations


# Regression

This is a key concept in scientific data analysis. We can briefly put the 
regression problem in general as looking for a function that "best explains"
a collection of response values as a function of some other feature values. 
This is also commonly referred to as "interpolation".

## Set-up

Suppose you have carried out an experiment leading to some table filled out 
with numerical values. As usual, each row represents a sample or replicate in 
our experiment; and each column represents the distinct measurements (features) 
conducted on each sample.

Let $Y$ be a distinguished feature (response variable) that we want to explain
as a function of another set of features: $X_1, \ldots, X_m$. 

In statistics, carrying out regression means to find the random process
that generated the data in $Y$ using as input the set of features. random
means that is subject to some level of randomness. Consequently, unlike the 
case of functions, we do not expect to get always same outputs from identical
inputs.

In general, this problem is very hard. In practice, to make life simple, 
we will search in a relatively small family of such processes (distributions). 

## Ingredients

We need to clarify a few things before daring to work on any 
regression problem at all. At least, we need to specify:

- The set of distributions (family) we will pick our solution from.
- A criterion to ascertain what a good solution is.

Also we will make a technical assumption, that implies that 
all samples in our dataset have been sampled independently
from the same random process.


### Distribution family

It is common that the distributions are taken from a family governed by
a few parameters:

- Gaussian family: $$f_{\mu, \sigma}(x)=\frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{1}{2} \left(\frac{x-\mu}{\sigma}\right)^2}$$ 
where $\mu$ is the mean and $\sigma$ is the standard deviation. For more information
check the [Normal distribution Wikipedia article](https://en.wikipedia.org/wiki/Normal_distribution).
- Bernoulli/Binomial family: $$g_p(x) = xp + (1-x)(1-p)$$ where $p$ is the probability of drawing a 1.
- Poisson family: $$h_{\lambda}(n) = \frac{\lambda^n}{n!}e^{-\lambda}$$ where $\lambda$ is the mean.

Then finding the right solution is equivalent to finding the **right parameters**.

### Likelihood: without covariates

For each distribution of choice we can compute something called **likelihood of 
the data given the distribution** or **likelihood** for short.
Given the response data $Y = \{y_1,...,y_n\}$, assuming the $y_i$ have been
produced by the random process encoded by the PMF or PDF $f_{\theta}$ (where $\theta$ 
represents a parameter that govern the distribution) the likelihood 
of the data is given by:

$$\mathcal{L}(\theta) = f_{\theta}(y_1)f_{\theta}(y_2)\ldots f_{\theta}(y_n) = \prod_{i=1}^n f_\theta(y_i)$$
We will be looking for the $\hat{\theta}$ that gives the maximum $\mathcal{L}(Y,\theta)$. 
In a more compact notation:
$$\hat{\theta} = \textrm{argmax}_{\theta} \; \mathcal{L}(\theta)$$
### Likelihood: with covariates

But what if we want to make the draw dependent on a set of covariates, the 
covariates we have been talking about before? In the context of generalized 
linear models, the trick is that they are hidden inside the parameters of 
the distribution.

### Example

Suppose our dataset has response vector $Y=(y_1, \ldots, y_n)$ and
a covariate vector $X=(x_1, \ldots, x_n)$. If we assume that $Y$ has been 
generated by distributions of a family governed by a parameter $\theta$, we 
will see $\theta$ as being dependent on $X$ in a linear fashion 
$\theta = \alpha X + \beta$. 

This has an important implication: the specific $\theta$ used to 
draw each $y_i$ (say $\theta_i$) depends on the covariate of this sample 
in a linear fashion: $\theta_i = \alpha x_i + \beta$. Now we want to find
$\hat\alpha$ and $\hat\beta$ such that:

$$ \mathcal{L} (\alpha, \beta) = f_{\theta_1}(y_1) f_{\theta_2}(y_2) \ldots f_{\theta_n}(y_n) = \prod_{i=1}^n f_{\theta_i}(y_i)$$
$$(\hat\alpha, \hat\beta) = \textrm{argmax}_{(\alpha, \beta)} \; \mathcal{L}(\alpha, \beta) $$

# Linear Regression

To illustrate linear regression from different angles, we will resort once more to 
the [gapminder dataset](https://cran.r-project.org/web/packages/gapminder/README.html).

This time we want to understand the statistical association between the GDP per capita 
and life expectancy across different countries for the year 2007.

Let's retrieve and preprocess our data:


```{r}
library(tidyverse)
gapminder <- as.data.frame(gapminder::gapminder)
gapminder_2002 <- gapminder %>% filter(year == 2007)
gapminder_sub <- gapminder_2002[, c("country", "lifeExp", "gdpPercap")]
gapminder_sub$log10_gdpPercap <- log10(gapminder_sub$gdpPercap)
head(gapminder_sub)
```

Recall that our goal is to compute a vector $\hat\beta=(\hat\beta_1, \hat\beta_2)$
such that if $A\in\mathbb{R}^{n\times m}$ is the matrix of covariates  across samples and $b\in\mathbb{R}^n$ is the vector of independent terms 
(response variable), $A\beta$ is a reasonable representation of what has been observed as $b$.

## Solution 1: orthogonal projection

We define the matrix of coefficients $A$ and the independent term $b$.

```{r}
num_rows <- nrow(gapminder_sub)
A <- matrix(
  c(gapminder_sub$log10_gdpPercap, rep(1, num_rows)),
  nrow = num_rows,
  ncol = 2,
  byrow = FALSE
)

b <- as.matrix(gapminder_sub$lifeExp)
```

Notice that the $A$ already incorporates a column of ones that renders an 
intercept after multiplying by the unknown vector of linear coefficients. 

We already know (and understand the geometrical meaning behind it, see [tutorial 
on orthogonal projections](https://koszulordie.github.io/mat/2020/11/04/session07.html))
at least one sensible way to proceed is to compute the solution $\hat{x}$ 
can be computed with the following recipe:

$$\hat{x} = (A^tA)^{-1} A^t b.$$

Let's do the computation:

```{r}
library(matlib)
xhat <- inv(t(A) %*% A) %*% t(A) %*% b
xhat
```

## Solution 2: GLM function in R

The most compact, straightforward (and lazy) way to solve this problem with R 
is by means of the functions available in R for GLM regression:

```{r}
glm(lifeExp ~ log10_gdpPercap + 1, family=gaussian, data=gapminder_sub)
```

There are already two remarkable things in this short expression:
- the formula "lifeExp ~ log10_gdpPercap + 1" employed as first argument
- the "family=gaussian" argument

## Solution 3: optimization (least squares)

Let's frame our problem as an optimization problem, based on minimizing the 
function that measures the sum of squares of the residuals:

### Objective function

Let's define the function that given the parameters of the model
(slope and intercept) gives the sum of squares of the residuals:


```{r}
squared_error <- function(x){
  v <- c(x[1], x[2]);
  C <- A%*%v - b;
  t(C) %*% C
}

# test the function at some point
a <- c(0,1)
squared_error(a)
```

What does the sum of squares of the residuals has to do with the previous 
discussion in the first place?

### How the square error arises from doing regression with the Gaussian family

The squared error arises naturally from the expression of the log-likelihood,
where the distributions employed are Gaussian:

$$f_{\mu, \sigma}(x)=\frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{1}{2} \left(\frac{x-\mu}{\sigma}\right)^2}$$
and the linear dependence with the covariates is incorporated to the $\mu$:

$$\mu_i = A_i\beta$$
with $A_i$ represents the $i$-th row of $A$, i.e., the covariate values
corresponding to the $i$-th sample.

Let's compute the likelihood function:

$$\mathcal{L}(\beta) = \prod_{i=1}^n f_{\mu_i, \sigma}(y_i) = \prod_{i=1}^n \frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{1}{2} \left(\frac{y_i-\mu_i}{\sigma}\right)^2}$$
Let's compute the log version:
$$
\begin{align}
\log\mathcal{L}(\beta) & = & -n\log\sqrt{2\pi\sigma^2} - \frac{1}{2\sigma^2}\sum_{i=1}^n (y_i-\mu_i)^2 \\ 
& \propto & \sum_{i=1}^n(y_i-A_i\beta)^2 = k - (A\beta - y)^t(A\beta - y)
\end{align}
$$
where $k$ is a constant terms which does not depend on $\beta$.

The only non-constant part of the expression involves the squared error function. 
The log-likelihood will be maximized if the squared error function is minimized.

### Gradient of the squared error

For a given function, there are packages that allow to compute their gradient at
some point numerically. For instance:

```{r}
library(rootSolve)
x0 <- c(15, 1)  # some example
grad <- gradient(squared_error, x0)
grad
```

However, interestingly, we can provide a closed form of the gradient 
of this specific objective function (see [Exercises for Session 13-14](https://koszulordie.github.io/mat/exercises/exercises-session-14.pdf)).

Let's implement it as a function and evaluate in the previously computed "xhat":

```{r}
grad_squared_error <- function(x){
  v <- c(x[1], x[2]);
  2 * t(A %*% v - b) %*% A
}
grad_squared_error(xhat)
squared_error(xhat)
```

We can use our gradient function implementation to find the root of the gradient 
numerically:

```{r}
library(rootSolve)
res <- multiroot(grad_squared_error, c(5,5), atol=c(1e-6,1e-6), maxiter=1e4)
xhat2 <- res$root
print(xhat2)
print(res$f.root)
```

### Graphical representation

To represent graphically the function we have just optimized, it is convenient
that we set-up a grid of inputs and add a new column with the outputs of our 
function:

```{r}
xs <- seq(10, 20, by=0.05)
ys <- seq(0, 10, by=0.05)

griddf <- expand.grid(alpha=xs, beta=ys)
griddf$value <- apply(griddf, 1, function(x) {squared_error(x)})
```

Let's visualize the function values around "xhat" as a 2D contour plot
filled with colors:

```{r}
se_gg <- griddf %>%
  ggplot(aes(alpha, beta, z=value)) +
  geom_contour_filled(bins=100, show.legend=FALSE) + 
  geom_point(aes(x=xhat[1], y=xhat[2]), colour="red") + 
  annotate("text", x=xhat[1], y=xhat[2], hjust = 1.2, vjust=1.2, label = "xhat", size = 5, color = "red") +
  coord_fixed(ratio = 1) + 
  labs(x="alpha", y="beta") + 
  ggtitle("Squared residuals function: colormap plot")

se_gg
```

Let's visualize the function using just the 2D level sets:

```{r}
se_gg <- griddf %>%
  ggplot(aes(alpha, beta, z=value)) +
  geom_contour(bins=1000, show.legend=FALSE) + 
  geom_point(aes(x=xhat[1], y=xhat[2]), colour="red") + 
  annotate("text", x=xhat[1], y=xhat[2], hjust = 1.2, vjust=1.2, label = "xhat", size = 5, color = "black") +
  coord_fixed(ratio = 1) +
  labs(x="alpha", y="beta") + 
  ggtitle("Squared residuals function: contour plots")

se_gg
```


Let's take advantage of our knowledge to compute the Hessian matrix (see also [Exercises for Session 13-14](https://koszulordie.github.io/mat/exercises/exercises-session-14.pdf))

```{r}
hessian <- 2 * t(A) %*% A
hessian
```

The eigenvalues and eigenvectors of the Hessian tell us about:

- **eigenvalues:** how open/closed the branches of the vertical cross-sections of the graph
of the function near "xhat" in the directions of the eigenvectors; these vertical 
cross-sections are all parabolas; and the graph of the function in this specific
example is known as an [elliptic parabolloid](https://en.wikipedia.org/wiki/Paraboloid#Elliptic_paraboloid).

- **eigenvectors:** orthogonal directions signaling the parabolic cross-sections
that have max/min branch openness.


```{r}
eigen <- eigen(hessian)

eigenvectors <- eigen$vectors
eigenvectors

eigenvalues <- eigen$values
eigenvalues
```

Let's see where those vectors are sitting in the contour plot.


```{r}
se_gg <- griddf %>%
  ggplot(aes(alpha, beta, z=value)) +
  geom_contour(bins=200, show.legend=FALSE) + 
  geom_point(aes(x=xhat[1], y=xhat[2]), colour="red") +
  geom_segment(aes(x=xhat[1], y=xhat[2], 
                   xend=xhat[1] + eigenvectors[1,1], yend=xhat[2] + eigenvectors[2,1]),
               arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x=xhat[1], y=xhat[2], 
                   xend=xhat[1] + eigenvectors[1,2], yend=xhat[2] + eigenvectors[2,2]),
               arrow = arrow(length = unit(0.2, "cm"))) + 
  coord_fixed(ratio = 1) + 
  annotate("text", x=xhat[1], y=xhat[2], hjust = 1.2, vjust=2.0, label = "xhat", size = 5, color = "black") +
  labs(x="alpha", y="beta") + 
  ggtitle("Squared residuals function: contour plots")

se_gg
```

# Logistic regression

To illustrate logistic regression we will resort to 
the [Breast Cancer Wisconsin Diagnostic Dataset from UCI Machine
Learning Repository](https://archive.ics.uci.edu/ml/datasets/breast+cancer+wisconsin+(diagnostic)) 
imported with **dslabs**.

```{r}
library(dslabs)
brca <- as.data.frame(dslabs::brca)
brca_sub <- brca[, c("x.radius_mean", "x.texture_mean", "y")]
brca_sub$y <- apply(brca_sub, 1, function(x) {x[3]=="M"})
```

We want to understand the statistical association between malignancy of the tumor
(yes or no) and a set of potential predictors derived from biopsies.


## GLM function in R


```{r}
res <- glm(y ~ x.radius_mean + x.texture_mean, family=binomial, data=brca_sub)
res$coefficients
```
## What's going on inside the blackbox?

In logistic regression we assume that the binary response $y_i$, 
is drawn from a Bernoulli with a probability $p_i$ that depends on the 
covariates.

More specifically, if $A\in\mathbb{R}^{n\times m}$ is 
the matrix that contains the values of the $m$ covariates for the $n$ 
samples, and $A_i$ denotes the $i$-th row, then

$$p_i = \frac{e^{A_i{\bf \beta}}}{1 + e^{A_i{\bf \beta}}}$$

where ${\beta} = (\beta_1,\ldots,\beta_m)$ is the vector of unknown 
parameters of the model. 

There is a helpful alternative way to stress this relationship:

$$\log\left(\frac{p_i}{1 - p_i}\right) = A_i{\beta}$$

What the function **glm** does boilds down to computing the parameter $\hat\beta$
that makes the following likelihood maximum:

$$
\begin{align}
\mathcal{L}(\beta) & = & \prod_{i=1}^n \left\{ y_ip_i + (1 - y_i)(1 - p_i)\right\} \\
& = & \prod_{i=1}^n \left\{ y_i \frac{e^{A_i{\bf \beta}}}{1 + e^{A_i{\bf \beta}}} + (1 - y_i)(1 - \frac{e^{A_i{\bf \beta}}}{1 + e^{A_i{\bf \beta}}})\right\}
\end{align}
$$
or its log version

$$
\begin{align}
\log\mathcal{L}(\beta) & = & \sum_{i=1}^n \log\left\{ y_ip_i + (1 - y_i)(1 - p_i)\right\} \\
& = & \sum_{i=1}^n y_i\log p_i + (1 - y_i)\log(1 - p_i).
\end{align}
$$

# Poisson regression

To illustrate Poisson regression we will resort to the 
[US Contagious Diseases](http://www.tycho.pitt.edu/) dataset imported with **dslabs**.
This dataset portrays yearly counts for Hepatitis A, Measles, Mumps, Pertussis, 
Polio, Rubella, and Smallpox for US states.


```{r}
library(dslabs)
diseases <- as.data.frame(dslabs::us_contagious_diseases)
diseases <- diseases %>% filter(state == "Arizona") %>% filter(disease == "Measles")
diseases$log10_population <- log10(diseases$population)
```

We would like to analyze the trends of measles cases reported in Arizona through 
time but would also like to understand the influence of the population size.

```{r}
res <- glm(count ~ log10_population, family=poisson, data=diseases)
res$coefficients
```


## What's going on inside the blackbox?

In Poisson regression we assume that the count response variable $n_i$, 
is drawn from a Poisson with mean parameter $\lambda_i$ that depends on the 
covariates. 

More specifically, if $A\in\mathbb{R}^{n\times m}$ is 
the matrix that contains the values of the $m$ covariates for the $n$ 
samples, and $A_i$ denotes the $i$-th row, then

$$\lambda_i = e^{A_i \beta}$$

where ${\beta} = (\beta_1,\ldots,\beta_m)$ is the vector of unknown 
parameters of the model. 

What the function **glm** does boilds down to computing the parameter 
$\hat\beta$ that makes the following likelihood maximum:

$$\mathcal{L}(\beta) = \prod_{i=1}^n \frac{\lambda^{n_i}_i}{n_i!}e^{-\lambda_i}$$
or its log version

$$
\begin{align}
\log\mathcal{L}(\beta) &=& \sum_{i=1}^n -\lambda_i + n_i\log\lambda_i - {n_i}! \\
&=& k + \sum_{i=1}^n -e^{A_i\beta} + n_i A_i\beta
\end{align}
$$
where $k$ is a constant terms which does not depend on $\beta$.

```{r}
sessionInfo()
```
